The Design of an Aliasing-Tolerant Parallel Programming Model

### 2012-11-08 doc/types-pldi2013/intro.tex
intro

Previous work has explored two approaches for describing the organization of data
in distributed memory programs.
One class of work uses fully static analyses with no runtime overhead, but 
disallows potential data aliasing to make the analysis tractable, limiting expressivity.
Two recent examples, Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
mechanism to statically partition the heap.
% into a tree of collections of data.
The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that must be checked statically 
(see Section~\ref{sec:related}).  The second approach permits potential data
aliasing, allowing greater expressivity, but uses fully dynamic analyses of
program execution to detect data aliasing and guarantee correct results.
This approach incurs significant runtime overhead and requires 
centralized control logic that is difficult to scale to distributed 
memory machines; examples include transactional memory \cite{Harris05} 
and thread-level speculation \cite{Steffan00}.

Our own experience writing high-performance applications
using the current industry standard mix of MPI, shared-memory
threads, and CUDA has taught us that a
single, static data partitioning is often insufficient.  In many cases, the
best way to partition data is often a function of the data
itself---i.e., the partitions must be dynamically computed.
Furthermore, applications often need multiple, simultaneous partitions of
the same data, each providing different views on to that data.

@@

To support additional parallelism in the presence of aliased data,
Legion provides {\em coherence} modes at the granularity of regions,
which allow relaxed constraints on task execution order, even
in cases where tasks access the same region.  The programmer may
specify that tasks conflicting on a region run: in
program order (the default), {\em atomically} with respect to each
other, or even {\em simultaneously} if the application has alternative
methods for ensuring determinism or can tolerate non-deterministic
behavior in that region.

@@

\item We present a parallel operational semantics for Core Legion that
  may be of independent interest.  This semantics is compositional,
  hierarchical, and asynchronous, reflecting the way such programs
  actually execute on the hardware.  The semantics also
  models conflicting memory operations between tasks in cases of relaxed
  coherence, making explicit the boundary between deterministic and
  possible non-deterministic behavior (Section~\ref{subsec:opsemantics}).

@@

\item We give experimental evidence that supports the Legion design choices.  On three
  real-world applications, we show that dynamic region pointer checks
  would be expensive, justifying checking this aspect of the type
  system statically.  We also show that the cost of region aliasing
  checks is low, showing that an expressive and dynamic
  language with aliasing is not incompatible with both high
  performance and safety (Section~\ref{sec:evaluation}).

>>>>

Previous work has explored two approaches for describing the organization of data
in distributed memory programs.
One class of work uses fully static analyses with no runtime overhead, but 
disallows potential data aliasing to make the analysis tractable, limiting expressivity.
Two recent examples, Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
mechanism to statically partition the heap.
% into a tree of collections of data.
The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that must be checked statically 
(see Section~\ref{sec:related}).  The second approach permits data
aliasing, allowing greater expressivity, but uses fully dynamic analyses of
program execution to detect aliasing and guarantee correct results.
This approach incurs significant runtime overhead and requires 
centralized control logic that is difficult to scale to distributed 
memory machines; examples include transactional memory \cite{Harris05} 
and thread-level speculation \cite{Steffan00}.

Our own experience writing high-performance applications
using the current industry standard mix of MPI, shared-memory
threads, and CUDA has taught us that a
single, static data partitioning is often insufficient.  In many cases, the
best way to partition data is often a function of the data
itself---i.e., the partitions must be dynamically computed and cannot be statically described.
Furthermore, applications often need multiple, simultaneous partitions of
the same data, which introduces aliasing.

@@

To support additional parallelism in the presence of aliased data,
Legion provides {\em coherence} modes at the granularity of regions,
which allow relaxed constraints on task execution order, even
in cases where tasks access the same region.  The programmer may
specify that tasks conflicting on a region run in
program order (the default), {\em atomically} with respect to each
other, or even {\em simultaneously} if the application has alternative
methods for ensuring determinism or can tolerate non-deterministic
behavior in that region.

@@

\item We present a novel parallel operational semantics for Core Legion.
 This semantics is compositional,
  hierarchical, and asynchronous, reflecting the way such programs
  actually execute on the hardware.  The semantics also
  models conflicting memory operations between tasks in cases of relaxed
  coherence, making explicit the boundary between deterministic and
  possible non-deterministic behavior (Section~\ref{subsec:opsemantics}).

@@

\item We give experimental evidence that supports the Legion design choices.  On three
  real-world applications, we show that dynamic region pointer checks
  would be expensive, justifying checking this aspect of the type
  system statically.  We also show that the cost of region aliasing
  checks is low, showing that an expressive and dynamic
  language with aliasing is compatible with both high
  performance and safety (Section~\ref{sec:evaluation}).
